### CS598 - PSL - CODING ASSIGNMENT 14
### Net ID: balajis2
### Date: 10/14/2020

#### STEP 0: Install required packages, library and set seed
```{r}
set.seed(8564) 
library(mclust)
```

#### E-step and M-step derivations


**E-step:**

The below derivations are used for calculating the new probabilities of the two components:

**For component 1**
$$
A = p1 * N(x;μ,Σ) / (p1 * N(x;μ,Σ) + p2 * N(x;μ,Σ))
$$
$$
N(x;μ,Σ) =  1/\sqrt{(2\pi)^2} * 1/\sqrt{|\sum|)} * exp^(-0.5 * (x-\mu)^t * (\sum)^-1 * (x-\mu))
$$

**For component 2**

$$
B = 1 - A
$$

**M-step**

The below equations for calculating the updated probabilities ($\hat{p1}, \hat{p2}$), mean ($\hat{\mu1},\hat{\mu2}$) and sigma ($\hat{\sum}$)


For $\hat{p1}, \hat{p2}$

$$\hat{p1} = \sum p1/n$$ 
$$\hat{p2} = \sum p2/n$$ 
                      
                      
For $\hat{\mu1},\hat{\mu2}$

$$\hat{\mu1} = \sum(\hat{p1} * x) / \sum(\hat{p1}) $$
$$\hat{\mu2} = \sum(\hat{p2} * x) / \sum(\hat{p2}) $$

For sigma $\hat{\sum}$

$$\hat{\sum} = 1/n * \sum(p1 * (X-\hat{\mu1}) + p2 * (X-\hat{\mu2}) $$

#### STEP 1: myEM function
```{r}

dbinorm<-function(x,m,s){
  return( 1/sqrt((2*pi)**2) * 1/sqrt(det(s)) * exp(-0.5 * t(x-m) %*% solve(s) %*% (x-m)) )
}

Estep <- function(data, G, para){
  # Your Code
  # Return the n-by-G probability matrix
  n <- nrow(data)
  new_prob_matrix <- matrix(0, n, G) 
  
  m1 = c(para$mean[1,1],para$mean[2,1])
  m2 = c(para$mean[1,2],para$mean[2,2])
  
  s = matrix(c(para$Sigma[1,1],para$Sigma[1,2],para$Sigma[2,1],para$Sigma[2,2]),2,2)
  p1 = para$prob[1]
  p2 = para$prob[2]
  
  for (i in 1:n){
    X = c(data[i,1],data[i,2])
    p = (p1 * dbinorm(X,m1,s)) / ( (p1 * dbinorm(X,m1,s)) + (p2*dbinorm(X,m2,s)))
    new_prob_matrix[i,1] = p
    new_prob_matrix[i,2] = 1-p
  }
  
  return(new_prob_matrix)
}

Mstep <- function(data, G, para, post.prob){ 
  
  # calculate updated probabilities
  new_p1 = mean(post.prob[,1])
  new_p2 = mean(post.prob[,2])
  
  # calculate updated probabilities
  new_m1 = (t(data)%*%diag(post.prob[,1])%*%rep(1,nrow(data)))/sum(post.prob[,1])
  new_m2 = (t(data)%*%diag(post.prob[,2])%*%rep(1,nrow(data)))/sum(post.prob[,2])
  
  # calculate updated sigma
  MY1 = matrix(new_m1,nrow(data),2,byrow=T)
  MY2 = matrix(new_m2,nrow(data),2,byrow=T)
  sigma = (t(data-MY1)%*%diag(post.prob[,1])%*%as.matrix(data-MY1)) + (t(data-MY2)%*%diag(post.prob[,2])%*%as.matrix(data-MY2))
  
  for (i in 1:2){
    sigma[i,1] = sigma[i,1]/nrow(data)
    sigma[i,2] = sigma[i,2]/nrow(data)
  }
  
  #Generate new parameters
  new_para = list(prob= cbind(as.matrix(new_p1),c(new_p2)), mean = cbind(as.matrix(new_m1),c(new_m2)), Sigma= sigma)
  
  #Return new parameters
  return(new_para)

  }

myEM <- function(data, itmax, G, para){
  # itmax: num of iterations
  # G:     num of components
  # para:  list of parameters (prob, mean, Sigma)
  
  for(t in 1:itmax){
    post.prob <- Estep(data, G, para)
    para <- Mstep(data, G, para, post.prob)
  }
  
  return(para)  
}
```
#### STEP 2: Load data
```{r}
n <- nrow(faithful)
Z <- matrix(0, n, 2) 
Z[sample(1:n, 120), 1] <- 1 
Z[, 2] <- 1 - Z[, 1]

ini1 <- mstep(modelName="EEE", faithful , Z)

ini0 <- mstep(modelName="EEE", faithful , Z)$parameters

para0 <- list(prob = ini0$pro, 
              mean = ini0$mean, 
              Sigma = ini0$variance$Sigma)

```
#### STEP 4: Test myEM function
```{r}
myEM(data=faithful, itmax=10, G=2, para=para0)
```
#### STEP 5: COMPARE OUTPUT FROM mcluster
```{r}
Rout <- em(modelName = "EEE", data = faithful,
           control = emControl(eps=0, tol=0, itmax = 10), 
           parameters = ini0)$parameters
list(Rout$pro, Rout$mean, Rout$variance$Sigma)
```